---
title: "StrataFinder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{StrataFinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StrataFinder)
```

# Load data
First, let's load some sample data. We will use data from Interochromis loock linkage group 5.
This data has F_{ST} counts for 36,170,306 base pairs. Counts for each 100,000 base pairs are grouped together.

```{r}
Load the data
data(example_data, package = "StrataFinder")
# df <- read.table("data/Interochromis_loock_LG5.txt", sep="\t")
head(example_data)
```

In order to predict the break points, we will use the index (base pair starting location divided by 100,000) as our $x$ values and the count as our $y$ values.

```{r}
# x <- as.integer(rownames(df))
# y <- as.integer(df$V4)
```

# Estimate breakpoints

Create a new StrataFinder object.

```{r}
sf <- new("StrataFinder")
```

To estimate the breakpoints, you will need to input the type of function you want to fit (either linear or exponential)
and the minimum length each segment can be. Here, we will fit a linear function with a minimum length of 10.

```{r}
fit_type <- "linear"
min_len = 10
```

If you already know how many breakpoints you need and would just like to estimate their location, you can use the findNStrata method.
```{r}
n_bp <- 2
# sf$findNStrata(x, y, n_bp, fit_type, min_len)
```

If you don't know how many breakpoints you need, you can estimate both the number of breakpoints and their locations using the findStrata method.
```{r}
min_bps <- 0
max_bps <- 2
# sf$findStrata(x, y, fit_type, min_len, min_bps, max_bps)
```

There are two parts to the algorithm behind StrataFinder: estimating the sum squared residual matrix and identifying the breakpoints that minimize the total error.
The methods findNStrata and findStrata will perform both steps.
If you have a large dataset or will be testing multiple number of breakpoints you want to evaluate, you can precompute the sum squared residual matrix and input it into findStrata or findNStrata.
This will reduce redundant computations and speed up run time.

```{r}
# ssr_mat <- sf$precomputeSSRMat(x, y, fit_type)
# sf$findNStrata(x, y, n_bp, fit_type, min_len, ssr_mat)
```